STA4241 – Final Project 
Flight Price Prediction Analysis
By Jack Griffin

Goals of this analysis:
 1) Explore flight price data and identify key drivers of price
 2) Model price or log(price) using multiple algorithms
 3) Compare model performance and select a final model
 
# STEP ONE: Setup -- Packages and Libraries
```{r}
# Run this install line to install packages, if necessary (uncomment it)
# install.packages(c("tidyverse", "readxl", "forcats", "caret", "glmnet", "ranger"))

library(tidyverse)   # ggplot2, dplyr, etc.
library(readxl)      # read_excel()
library(forcats)     # factor level utilities (fct_infreq, fct_relevel)
library(caret)       # createDataPartition() for stratified split
library(glmnet)      # ridge / lasso / elastic net
library(ranger)      # faster version of random forest
```

# STEP TWO: Load & Check Dataset
```{r}
# Load Excel dataset
flights <- read_excel("Flight_Project_Dataset.xlsx")

# Basic structure checks
dim(flights)        # number of rows and columns
names(flights)      # variable names
glimpse(flights)    # types and first few entries
summary(flights)    # quick descriptive statistics

# Missing values and duplicates
sum(is.na(flights))          # total number of NAs
colSums(is.na(flights))      # NAs by column
sum(duplicated(flights))     # number of duplicate rows
```

# STEP THREE: Explore Numeric Variables
```{r}
# Histogram of raw price
ggplot(flights, aes(x = price)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(title = "Flight Price Distribution",
       x = "Price", y = "Count")

# Histogram of log(price) to check skewness
ggplot(flights, aes(x = log(price))) +
  geom_histogram(bins = 40, fill = "darkorange", color = "white") +
  labs(title = "Log-Transformed Flight Price Distribution",
       x = "log(Price)", y = "Count")

# Correlation matrix for numeric predictors
num_vars   <- flights %>% select(price, duration, days_left)
cor_matrix <- cor(num_vars)
cor_matrix
```

# STEP FOUR: Log Transform Price & Inspect Categorical Effects
```{r}
# Create transformed target variable
flights <- flights %>%
  mutate(log_price = log(price))

# Boxplots: raw price vs class and stops
ggplot(flights, aes(x = class, y = price)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Price by Class", x = "Class", y = "Price (₹)")

ggplot(flights, aes(x = stops, y = price)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Price by Number of Stops", x = "Stops", y = "Price (₹)")

# Boxplots: log(price) vs class and stops
ggplot(flights, aes(x = class, y = log_price)) +
  geom_boxplot() +
  labs(title = "log(Price) by Class", x = "Class", y = "log(Price)")

ggplot(flights, aes(x = stops, y = log_price)) +
  geom_boxplot() +
  labs(title = "log(Price) by Number of Stops", x = "Stops", y = "log(Price)")

# Grouped summaries (mean / median price and log_price)
by_class <- flights %>%
  group_by(class) %>%
  summarise(
    n              = n(),
    mean_price     = mean(price),
    median_price   = median(price),
    mean_log_price = mean(log_price),
    .groups        = "drop"
  )

by_stops <- flights %>%
  group_by(stops) %>%
  summarise(
    n              = n(),
    mean_price     = mean(price),
    median_price   = median(price),
    mean_log_price = mean(log_price),
    .groups        = "drop"
  )

by_class
by_stops
```

# STEP FIVE: Train/Test Split on Stratified Log(Price)
```{r}
set.seed(4241)

# Stratified split to keep log_price distribution similar in train and test
idx   <- createDataPartition(flights$log_price, p = 0.75, list = FALSE)
train <- flights[idx, ]
test  <- flights[-idx, ]

# Check train/test sizes and similarity of log_price and price
nrow(train); nrow(test)

train_summ <- train %>%
  summarise(mean_log = mean(log_price),
            sd_log   = sd(log_price),
            mean_price = mean(price),
            sd_price   = sd(price))

test_summ <- test %>%
  summarise(mean_log = mean(log_price),
            sd_log   = sd(log_price),
            mean_price = mean(price),
            sd_price   = sd(price))

train_summ
test_summ

# Compare distributions of key categorical variables
prop_train_class <- prop.table(table(train$class))
prop_test_class  <- prop.table(table(test$class))

prop_train_stops <- prop.table(table(train$stops))
prop_test_stops  <- prop.table(table(test$stops))

prop_train_class; prop_test_class
prop_train_stops; prop_test_stops
```

# STEP SIX: Baseline Multiple Linear Regression on Log(Price)
```{r}
# Set reference levels for factors (most frequent categories)
train <- train %>%
  mutate(
    class          = fct_infreq(class) %>% fct_relevel("Economy"),
    stops          = fct_infreq(stops) %>% fct_relevel("one"),
    airline        = fct_infreq(airline) %>% fct_relevel(levels(fct_infreq(airline))[1]),
    source_city    = fct_infreq(source_city) %>% fct_relevel(levels(fct_infreq(source_city))[1]),
    destination_city = fct_infreq(destination_city) %>%
                       fct_relevel(levels(fct_infreq(destination_city))[1]),
    departure_time = fct_infreq(departure_time) %>%
                     fct_relevel(levels(fct_infreq(departure_time))[1]),
    arrival_time   = fct_infreq(arrival_time) %>%
                     fct_relevel(levels(fct_infreq(arrival_time))[1])
  )

# Apply the same levels to test data (prevents new-level issues)
test <- test %>%
  mutate(
    class          = factor(class,          levels = levels(train$class)),
    stops          = factor(stops,          levels = levels(train$stops)),
    airline        = factor(airline,        levels = levels(train$airline)),
    source_city    = factor(source_city,    levels = levels(train$source_city)),
    destination_city = factor(destination_city,
                               levels = levels(train$destination_city)),
    departure_time = factor(departure_time, levels = levels(train$departure_time)),
    arrival_time   = factor(arrival_time,   levels = levels(train$arrival_time))
  )

# Fit baseline multiple linear regression model
mlr_full <- lm(
  log_price ~ class + stops + duration + days_left +
    airline + source_city + destination_city +
    departure_time + arrival_time,
  data = train
)

summary(mlr_full)
```

# STEP SEVEN: Baseline MLR Performance (Training Set)
```{r}
# Predictions on training data (log and original price scales)
train$mlr_pred_log   <- predict(mlr_full, newdata = train)
train$mlr_pred_price <- exp(train$mlr_pred_log)

# RMSE and R-squared on log scale
postResample(train$mlr_pred_log, train$log_price)

# RMSE on original price scale
rmse_price_train <- sqrt(mean((train$price - train$mlr_pred_price)^2))
rmse_price_train
```

# STEP EIGHT: Regularized Regression (Ridge/Lasso/Elastic Net)
```{r}
# Use a smaller training subset for glmnet to reduce computation time
set.seed(4241)
train_small <- train %>% sample_n(25000)

# Create model matrix (glmnet requires numeric matrix; factors are one-hot encoded)
x_train <- model.matrix(
  log_price ~ class + stops + duration + days_left +
    airline + source_city + destination_city +
    departure_time + arrival_time,
  data = train_small
)[, -1]   # drop intercept column

y_train <- train_small$log_price

# Fit ridge, elastic net, and lasso models via cross-validated glmnet
alphas <- c(0, 0.5, 1)     # 0 = ridge, 0.5 = elastic net, 1 = lasso
names(alphas) <- c("ridge", "elastic_net", "lasso")

glmnet_models <- lapply(alphas, function(a) {
  cv.glmnet(
    x      = x_train,
    y      = y_train,
    alpha  = a,
    nfolds = 5     # 5-fold CV for speed
  )
})

# Compare CV errors and choose best alpha
cv_errors <- sapply(glmnet_models, function(m) min(m$cvm))
cv_errors

best_name  <- names(which.min(cv_errors))   # "lasso" in this run
best_model <- glmnet_models[[best_name]]
best_name

# Predict on FULL test set with the selected glmnet model
x_test <- model.matrix(
  log_price ~ class + stops + duration + days_left +
    airline + source_city + destination_city +
    departure_time + arrival_time,
  data = test
)[, -1]

test$glmnet_pred_log   <- as.numeric(predict(best_model, newx = x_test, s = "lambda.min"))
test$glmnet_pred_price <- exp(test$glmnet_pred_log)

# Coefficients at best lambda (for interpretation)
glmnet_coefs <- coef(best_model, s = "lambda.min")
glmnet_coefs
```

# STEP NINE: Tree-Based Model -- Random Forest via ranger
```{r}
# Fit random forest on the same 25k subset for efficiency
set.seed(4241)

rf_model <- ranger(
  formula    = log_price ~ class + stops + duration + days_left +
                             airline + source_city + destination_city +
                             departure_time + arrival_time,
  data       = train_small,
  num.trees  = 200,       # number of trees
  mtry       = 4,         # predictors tried at each split
  importance = "impurity"
)

# Predict on full test set
rf_pred <- predict(rf_model, data = test)
test$rf_pred_log   <- rf_pred$predictions
test$rf_pred_price <- exp(test$rf_pred_log)

# Variable importance (top 10)
rf_vi <- sort(rf_model$variable.importance, decreasing = TRUE)
rf_vi

barplot(rf_vi[1:10],
        las = 2,
        main = "Random Forest Variable Importance (Top 10)")
```

# STEP TEN: Final Test-Set Model Comparison
```{r}
# Fitting Random Forest predictions
rf_pred <- predict(rf_model, data = test)
test$rf_pred_log   <- rf_pred$predictions
test$rf_pred_price <- exp(test$rf_pred_log)

# Baseline MLR predictions on TEST set
test$mlr_pred_log   <- predict(mlr_full, newdata = test)
test$mlr_pred_price <- exp(test$mlr_pred_log)

# Helper functions for metrics
rmse <- function(obs, pred) {
  sqrt(mean((obs - pred)^2))
}

rsq <- function(obs, pred) {
  cor(obs, pred)^2
}

# Build comparison table for all three models
model_compare <- data.frame(
  Model = c(
    "Baseline MLR",
    "Elastic Net (alpha = 0.5)",     # best_name in this run
    "Random Forest (ranger)"
  ),
  
  RMSE_log = c(
    rmse(test$log_price, test$mlr_pred_log),
    rmse(test$log_price, test$glmnet_pred_log),
    rmse(test$log_price, test$rf_pred_log)
  ),
  
  R2_log = c(
    rsq(test$log_price, test$mlr_pred_log),
    rsq(test$log_price, test$glmnet_pred_log),
    rsq(test$log_price, test$rf_pred_log)
  ),
  
  RMSE_price = c(
    rmse(test$price, test$mlr_pred_price),
    rmse(test$price, test$glmnet_pred_price),
    rmse(test$price, test$rf_pred_price)
  )
)

model_compare
```